---
title: "02 QC and Preliminary Analysis"
author: "Jussi Kupari"
date: "Last edited: `r Sys.Date()`"
output: html_notebook
---

```{r}
# Load libraries and custom functions
suppressPackageStartupMessages({
  library(tidyverse)
  library(magrittr)
  library(Seurat)
  library(harmony)
  library(scPred)
})

R.utils::sourceDirectory(here::here("R", "functions"), modifiedOnly = FALSE)
```

```{r}
# Load preprocessed data
(drg_ra <- readRDS(here::here("data", "proc", "drg_ra_preprocessed.rds")))
```

```{r, fig.width=12, message=FALSE}
# First check with violins
VlnPlot(
  drg_ra,
  c("nCount_RNA", "nFeature_RNA", "percent.mt"),
  pt.size = 0,
  group.by = "orig.ident"
)
```

```{r, fig.width=12, message=FALSE}
# Detected gene number distribution
drg_ra@meta.data %>% 
  ggplot(aes(nFeature_RNA, fill = orig.ident)) +
  geom_histogram(alpha = 0.7, color = "white") +
  geom_vline(xintercept = 2000, color = "red") +
  geom_vline(xintercept = 1000, color = "blue") +
  facet_wrap(vars(orig.ident), scales = "free") +
  cowplot::theme_cowplot() +
  scale_x_continuous(labels = scales::label_number(suffix = " K", scale = 1e-3)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  theme(legend.position = "none") 
```

#### Create two datasets with cutoffs at 1000 and 2000 detected features
```{r}
# QC filter for > 2000 Features and percent.mt < 20
drg_ra_2000 <-
  drg_ra %>% 
  subset(percent.mt < 20 & nFeature_RNA > 2000)
```

```{r}
# QC filter for 999 > Features < 2001 and and percent.mt < 20
drg_ra_1000 <-
  drg_ra %>% 
  subset(percent.mt < 20 & nFeature_RNA < 2001) %>% 
  subset(nFeature_RNA > 999)
```

#### Clustering and celltype identification with > 2000 genes as cutoff
```{r}
# Integrate with harmony
drg_ra_2000 %<>% run_harmony()
```

```{r}
# Cluster cells
drg_ra_2000 %<>% cluster_seurat(reduction = "harmony", dims = 1:30, resolution = 0.1)
```

```{r, fig.width=12}
# Plot UMAPS
(DimPlot(drg_ra_2000, label = TRUE, shuffle = TRUE) + NoLegend()) +
  (DimPlot(drg_ra_2000, shuffle = TRUE, group.by = "orig.ident") + NoLegend()) +
  (DimPlot(drg_ra_2000, group.by = "timepoint_days", shuffle = TRUE) ) 
```

```{r, fig.width=12}
# Plot split UMAPs by timepoint
DimPlot(drg_ra_2000, label = TRUE, shuffle = TRUE, split.by = "timepoint_days", ncol = 4) + NoLegend() 
```

```{r, fig.width=12}
# See where the neurons are
FeaturePlot(drg_ra_2000, "Rbfox3", label = TRUE)
```

```{r}
# Add neuron identifier
drg_ra_2000@meta.data %<>% 
  mutate(is_neuron = seurat_clusters %in% c(3, 5, 7, 9, 12))
```

```{r}
# Save integrated full object
saveRDS(drg_ra_2000, file = here::here("data", "proc", "drg_ra_full_first_integration_2000genes.rds"))
```

#### Identify nonneuronal celltypes
```{r}
# Subset nonneurons
(nonneurons_2000 <-
  drg_ra_2000 %>% 
  set_idents("is_neuron") %>% 
  subset(idents = "FALSE"))
```

```{r}
# Use control cells only to identify celltypes
(nonneurons_ctrl_2000 <- 
  nonneurons_2000 %>% 
  set_idents("timepoint_days") %>% 
  subset(idents = "0"))
```

```{r}
# Run Harmony
nonneurons_ctrl_2000 %<>% run_harmony()
```

```{r}
# Cluster nonneurons
nonneurons_ctrl_2000 %<>% cluster_seurat(reduction = "harmony", dims = 1:20, resolution = 1)
```

```{r, fig.width=12}
# Plot UMAPS
(DimPlot(nonneurons_ctrl_2000, group.by = "orig.ident", shuffle = TRUE) + NoLegend()) +
  (DimPlot(nonneurons_ctrl_2000, label = TRUE, shuffle = TRUE) + NoLegend()) 
```

```{r}
# Add manual celltypes for non-neurons based on markers from Yim et al.
# https://doi.org/10.1038/s41593-021-01005-1
nonneurons_ctrl_2000@meta.data %<>% 
  mutate(cell_type = case_when(
    seurat_clusters %in% c(0, 1, 2, 5, 6) ~ "SatelliteGlia",
    seurat_clusters %in% c(4, 7, 21)      ~ "EndoNeuralCell",
    seurat_clusters %in% c(10, 16)        ~ "MyelSchwannCell",
    seurat_clusters == 20                 ~ "Pericyte",
    seurat_clusters == 15                 ~ "VSMC",
    seurat_clusters %in% c(8, 17, 14)     ~ "ImmuneCell",
    seurat_clusters == 9                  ~ "NonMyelSchwannCell",
    seurat_clusters == 13                 ~ "LymphaticEndothelialCell",
    seurat_clusters %in% c(3, 22, 18, 19) ~ "MicrovascularEndothelialCell",
    seurat_clusters == 12                 ~ "PerineurialCell",
    seurat_clusters == 11                 ~ "EpineurialCell",
    seurat_clusters == 5                  ~ "MyelSchwannCell"
  ))
```

```{r, fig.width=12}
# Plot UMAPS with celltypes
(DimPlot(nonneurons_ctrl_2000, group.by = "cell_type", label = TRUE, shuffle = TRUE, repel = TRUE))
```

```{r}
# Transfer labels to all nonneurons
nonneurons_ctrl_2000 %<>% set_idents("cell_type")
nonneurons_2000 %<>% predict_labels(reference = nonneurons_ctrl_2000, dims = 1:20)
```

```{r}
# Update full data with high level cell type slot (neuron or one of nonneuronal types)
drg_ra_2000@meta.data %<>% 
  rownames_to_column("cell") %>% 
  left_join(
    nonneurons_2000@meta.data %>%
      rownames_to_column("cell") %>% 
      select(cell, predicted.id),
    by = c("cell")
  ) %>% 
  replace_na(list(predicted.id = "Neuron")) %>% 
  rename(cell_class = predicted.id) %>% 
  column_to_rownames("cell")
```

```{r, fig.width=12}
# Plot UMAPS with high level cell types
DimPlot(drg_ra_2000, group.by = "timepoint_days", shuffle = TRUE) + 
(DimPlot(drg_ra_2000, group.by = "cell_class", label = TRUE, shuffle = TRUE, repel = TRUE) + NoLegend())
```

```{r}
# Save object with high level cell types
saveRDS(drg_ra_2000, file = here::here("data", "proc", "all_cells_after_first_clustering_2000genes.rds"))
```

#### Collect remaining non-neuronal cells that have 999 < Features < 2001
```{r}
# Integrate with harmony
drg_ra_1000 %<>% run_harmony()
```

```{r}
# Cluster cells
drg_ra_1000 %<>% cluster_seurat(reduction = "harmony", dims = 1:30, resolution = 0.1)
```

```{r, fig.width=12}
# Plot UMAPS
(DimPlot(drg_ra_1000, label = TRUE, shuffle = TRUE) + NoLegend()) +
  (DimPlot(drg_ra_1000, shuffle = TRUE, group.by = "orig.ident") + NoLegend()) +
  (DimPlot(drg_ra_1000, group.by = "timepoint_days", shuffle = TRUE) ) 
```

```{r, fig.width=12}
# Plot split UMAPs by timepoint
DimPlot(drg_ra_1000, label = TRUE, shuffle = TRUE, split.by = "timepoint_days", ncol = 4) + NoLegend() 
```

```{r, fig.width=12}
# Check where the neurons are
FeaturePlot(drg_ra_1000, c("Rbfox3", "Slc17a6"), label = TRUE)
```

```{r}
# Transfer labels from 2000 feature dataset
drg_ra_2000 %<>% set_idents("cell_class")
drg_ra_1000 %<>% predict_labels(reference = drg_ra_2000, dims = 1:30)
```

```{r, fig.width=12}
DimPlot(drg_ra_1000, label = TRUE, group.by = "predicted.id", repel = TRUE) 
```

```{r}
# Add a cutoff for prediction score
drg_ra_1000@meta.data %<>% 
  mutate(predicted_id = ifelse(prediction.score.max < 0.55, "unassigned", predicted.id))
```

```{r, fig.width=12}
# Plot with cutoff
DimPlot(drg_ra_1000, label = TRUE, group.by = "predicted_id", repel = TRUE) 
```

```{r}
# Save full drg_ra_1000 obejct
saveRDS(drg_ra_1000, file = here::here("data", "proc", "all_cells_after_first_clustering_1000genes.rds"))
```

```{r}
# Filter out unassigned cells
(drg_ra_1000_only_predicted <-
  drg_ra_1000 %>%
  set_idents("predicted_id") %>%
  subset(
    idents = drg_ra_1000@meta.data$predicted_id %>% unique() %>% keep(~ .x != "unassigned")
  ))
```

```{r, fig.width=12}
# Plot filtered secondary dataset
DimPlot(drg_ra_1000_only_predicted, label = TRUE, group.by = "predicted_id", repel = TRUE) 
```

```{r}
# Save filtered secondary dataset
saveRDS(drg_ra_1000_only_predicted, file = here::here("data", "proc", "all_cells_after_first_clustering_1000genes_only_predicted.rds"))
```

#### Merge secondary data non-neurons with primary data
```{r}
# Merge secondary data non-neurons with primary data
(drg_ra_full <-
  drg_ra_1000_only_predicted %>% 
  subset(
    idents = 
      drg_ra_1000_only_predicted@meta.data$predicted_id %>% 
      unique() %>% 
      keep(~ .x != "Neuron")
  ) %>% 
  merge(drg_ra_2000))
```

```{r}
# Fix cell_class in metadata
drg_ra_full@meta.data %<>% 
  mutate(cell_class = ifelse(is.na(predicted_id), cell_class, predicted_id))
```

```{r}
# Run harmony and cluster
drg_ra_full %<>% 
  run_harmony() %>% 
  cluster_seurat(dims = 1:30, resolution = 0.1) %>% 
  set_idents("cell_class")
```

```{r, fig.width=12, results='hide'}
# Plot UMAP of final merged dataset
DimPlot(drg_ra_full, label = TRUE, repel = TRUE) 
```

```{r, fig.width=12}
# Plot DotPlot for celltype markers from Yim et al. (and basic sensory neuron markers)
# https://doi.org/10.1038/s41593-021-01005-1
drg_ra_full %>%
  DotPlot(
    features =
      c("Pdgfra",
        "Egfr",
        "Cd74",
        "Mrc1",
        "Rbpj",
        "Ptprc",
        "Prx",
        "Scn7a",
        "Cnp",
        "Plp1",
        "Fabp7",
        "Ncam1",
        "Slc35f1",
        "Acta2",
        "Itgbl1",
        "Pdgfrb",
        "Cldn5",
        "Slit2",
        "Prox1",
        "Lyve1",
        "Opcml",
        "Rbfox3",
        "Slc17a6")
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r, fig.width=12}
# Plot celltype proportions in each timepoint
drg_ra_full@meta.data %>% 
  count(timepoint_days, cell_class) %>% 
  group_by(timepoint_days) %>% 
  mutate(pct = prop.table(n) * 100) %>%
  ggplot(aes(factor(timepoint_days), pct, fill = cell_class)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label=paste0(sprintf("%1.1f", pct),"%")),
            position=position_stack(vjust=0.5)) +
  labs(x = "Timepoint in days", y = "Proportion of cells", fill = "Cell class") +
  cowplot::theme_cowplot() +
  scale_y_continuous(expand = expansion(mult = c(0, .1)))
```

```{r}
# Save full integrated data
saveRDS(drg_ra_full, file = here::here("data", "proc", "drg_ra_full_merged.rds"))
```

```{r}
# Sessioninfo
sessionInfo()
```

```{r}
# Done!
clear_libraries()
```